---
title: "Penguins Data R Analysis"
author: "Niroshan Benjamin"
date: 2026-09-23
---

Data: [Palmer Penguins](https://allisonhorst.github.io/palmerpenguins/), Palmer
Station Antarctica LTER. The data ships with the `palmerpenguins` R package under a
[CC0 licence](https://creativecommons.org/publicdomain/zero/1.0/).

Cited as the package asks: Horst AM, Hill AP, Gorman KB (2020). *palmerpenguins:
Palmer Archipelago (Antarctica) penguin data*. R package version 0.1.1.
<https://allisonhorst.github.io/palmerpenguins/>.
doi: [10.5281/zenodo.3960218](https://doi.org/10.5281/zenodo.3960218).

The ggplot code in this post is adapted from work I did in DSCI 523 Lab 2, where the
plotting template was provided as part of the lab. 

Two of the penguins are missing their bill and body mass measurements, so they
are dropped from the figures below.

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(palmerpenguins)
```

```{r}
#| label: species-summary

pengSummary <- penguins |>
  group_by(species) |>
  summarise(count = n(),
            median_bill_length_mm = median(bill_length_mm, na.rm = TRUE),
            median_body_mass_g = median(body_mass_g, na.rm = TRUE))

pengSummary
```

```{r}
#| label: arrange-data

pengArr <- penguins |>
  arrange(desc(bill_length_mm))

head(pengArr)
```

```{r}
#| label: fig-arranged
#| fig-cap: "Bill length against body mass, with species in the default alphabetical legend order."
#| warning: false

penguins_originalArr <- ggplot(data = pengArr, mapping = aes(x = bill_length_mm, 
                                                           y = body_mass_g,
                                                           colour = species)) +
  geom_point(size = 2) +
  xlab("Bill length (mm)") +
  ylab("Body mass (g)") +
  theme_bw() +
  theme(text = element_text(size = 18))
penguins_originalArr
```

```{r}
#| label: reorder-factor

pengFct  <- penguins |>
  mutate(species = fct_rev(fct_reorder(species, body_mass_g, .na_rm = TRUE)))

levels(pengFct$species)
```

```{r}
#| label: fig-reordered
#| fig-cap: "The same data after reversing the species factor levels, which changes the legend order and the colours."
#| warning: false

penguins_originalFct <- ggplot(data = pengFct, mapping = aes(x = bill_length_mm, 
                                                             y = body_mass_g,
                                                             colour = species)) +
  geom_point(size = 2) +
  xlab("Bill length (mm)") +
  ylab("Body mass (g)") +
  theme_bw() +
  theme(text = element_text(size = 18))
penguins_originalFct
```

## Findings

Using group_by and summarise, I found the count and the median 'bill_length_mm' and
'body_mass_g' for each species. 'Gentoo' had the highest median body mass at 5000 g,
while 'Adelie' and 'Chinstrap' had the exact same median body mass of 3700 g.
'Chinstrap' had the longest median bill length at 49.6 mm, which is longer than
'Gentoo' even though 'Gentoo' is much heavier, so body mass and bill length do not
increase together across species.

Using the dplyr arrange function, I arranged the column 'bill_length_mm' in
descending order, but it had no effect on the figure. I used fct_reorder to reorder
the species based on 'body_mass_g', which also had no effect on the figure, so I
used fct_rev, which reversed the species order. This led to the figure changing,
with 'Gentoo' being the first species in the legend and having a red plot colour
now. The data points on the plot remained the same, but the colour of the points
changed. This is because factor levels are ordered alphabetically by default, and we
can see that changing levels impacts analysis and visualization.